Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-04-12 paper_jasa.tex
typos

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics.
%These plots should be familiar, because they are all in common use. 
For example, to examine the effect of variable $X_j$ on $Y$, we would plot residuals obtained from fitting the model without $X_j$ against $X_j$ or for a single covariate we may plot $Y$ against $X_j$ (cases 1-4 in Table~\ref{tbl:stat_multiple}). To assess whether the linear assumption is appropriate we would draw a plot of residuals against fitted values (case 5 in Table~\ref{tbl:stat_multiple}). For the purpose of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable, respectively.

@@

\begin{table*}[hbtp] 
\caption{Visual test statistics for testing hypotheses related to the model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2cm}m{3cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}\end{minipage} & Scatter plot with least square line overlaid. For null plots   we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Residual plot & \begin{minipage}[t]{3cm}   \scalebox{0.4}{\includegraphics{stat_beta_k.pdf}}\end{minipage} & Residual vs $X_k$ plots. For null plots   we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
3 & $H_0: \beta_k=0$ (for binary $X_k$) & Box plot & \begin{minipage}[t]{3cm} 	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For null plots   we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction of continuous and  binary $X_k$) & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_interection.pdf}}  \end{minipage} & Scatter plot with least square lines of each category overlaid. For null plots   we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_nonlinear.pdf}}  \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For null plots   we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{3cm} \scalebox{0.4}{\includegraphics{stat_sigma_box.pdf}}\end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For null plots   we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot of Residuals obtained by fitting partial regression. For null plots   we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_goodness_simple.pdf}}  \end{minipage} & Histogram of the response data. For null plots   we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & Special case $p=1$ $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot with least square line overlaid For null plots   we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

@@

\begin{equation}\label{power_exp} 
%   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$  increases, and when $\beta$ gets larger.   Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. We observe this dichotomy of visual power  in the data collected from user experiments, too. It features prominently in  figure \ref{fig:power_loess_effect}.

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  Figure \ref{fig:test_category} is an example of one of these lineups. It was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. The actual data plot location is ($4^2-1$). For this lineup, 15 out of 16 observers picked the actual data plot.

@@

% Entering 1st row
&  5 & 0, 1,  3, 5, 8  & 0.25, 0.75, 1.25, 1.75, 2.75 & 0.1, 0.4, 0.75, 1.25, 1.5, 2.25\\[-1ex]
\raisebox{1.5ex}{100} &12
& 1, 3, 8, 10, 16  & 0.5, 1.5, 3.5, 4.5, 6 &\\[1ex]

% Entering 2rd row
&  5 & 0, 1, 2, 3, 5  & 0.1, 0.4, 0.7, 1, 1.5&\\[-1ex]
\raisebox{1.5ex}{300} & 12
& 1, 3, 5, 7, 10  & 0, 0.8, 1.75, 2.3, 3.5 &\\[1ex]
% [1ex] adds vertical space
\hline
% inserts single-line
\end{tabular}
}
\label{tbl:experiment_params}
\end{table}

@@

This experiment is very similar to the previous one, except that there is a single continuous covariate and no second covariate (Equation \ref{multi} with $p=1$), following the test in Table \ref{tbl:stat_multiple}, case 2. Data is simulated with two sample sizes ($n=100, 300$), two standard deviations of the error ($\sigma=5, 12$), a variety of slopes ($\beta$), as given in Table  \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ and values for $X_1$ are simulated from $N(0,1)$. For each combination of parameters, at least three different actual data sets are produced, yielding a total of 70 lineups.

The actual data plot is generated by making a scatterplot of $Y$ vs $X_1$ with the least squares regression line overlaid. To produce the null plots in the lineup null data was simulated from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and plotted using the same scatterplot method as the actual data. 
To select 10 lineups for a subject, each combination of sample size ($n$) and error SD ($\sigma$) is given a difficulty value based on the slope ($\beta$) parameters. For the smallest slopes the difficulty is 4 (hardest) and for the largest slopes the difficulty is 0 (easiest). 
Figure \ref{fig:test_continuous} shows an example lineup for this experiment from difficulty level 4.  This lineup is generated using sample size ($n$)=100, slope($\beta$)=1.25 and and error SD($\sigma$)=5. The actual data plot location is $(2^2+1)$. None of the 65 observers picked the actual plot while 46 observers picked plot 18 which has the lowest $p$-value among all the plots in this lineup.

For each combination of sample size and standard deviation, the subject sees a randomly selected lineup of each difficulty level, giving 5 lineups. Another four lineups are chosen from a second choice of sample size and standard deviation, with difficulty levels 0-3, and the last lineup shown was a randomly selected lineup with difficulty 0. The order that the lineups are shown to participants is randomized. Each subject saw a substantial number of difficult lineups.

@@

The first two simulation experiments use data generated under a normal error model, satisfying the conditions for conventional test procedures. In these situations there exists a test, and there would, in general, be no need to use visual inference. The simulation is conducted with the hopes that the visual test procedure,  will at least compare favorably with the conventional test -- without any ambition of performing equally well.
 This third simulation is closer to the mark for the purpose of visual inference. The assumptions for the conventional test are violated, by contaminating the data. The contamination makes the estimated slopes effectively 0, yet the true value of slope parameter is not.  The data is generated from the following model:

@@

%{\bf Screening scenarios:}
Amazon Mechanical Turk workers are paid for their efforts, not substantially, but on the scale of the minimum wage of the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we take action based on a subject's answer to one of the very easy lineups that everybody was exposed to:
if the subject failed to identify the actual data plot, we remove all of this subject's data from the analysis. If the answer on the lineup is correct, we remove this lineup from the analysis, but keep all of the remaining answers. Table \ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.  A summary of all the data along with cleaned data are added to the supplementary material.

@@

\subsection{Model fitting}
 For each parameter combination, the effect is computed as 
\[
E=\sqrt {n} \cdot \beta/\sigma.
\]
\noindent
The model in Equation \ref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept, i.e. 
 $W_{\ell i} = E_{\ell i}$. Instead of fitting an intercept, we make use of a fixed  offset of $\log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different skill levels of subjects are accounted for  by allowing subject-specific random  slopes for  effect ($E$).

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.55}{\includegraphics{power_loess_effect_k.pdf}}
       \caption{Power in comparison to effect for the three experiments. Points indicate subject responses, with size indicating count. Responses are 1 and 0 depending on the success or failure respectively to identify the actual plot in the lineup. The loess curve (continuous line) estimates the observed proportion correct (power for $K=1$), and surrounding bands show simultaneous bootstrap confidence band. Observed proportion is used to obtain power for $K=5$. Conventional test power is drawn as a dashed line. For experiment 3, conventional power is based on the slopes of the non-contaminated part of the data. Power for contaminated data is shown by cross mark.}
       \label{fig:power_loess_effect}
\end{figure}

@@

Results of experiment 3 are curious insofar, as  power of the visual test is  high independent of effect size. However, these results are based on correct identification of the actual data plot, regardless of reason. Although subjects were asked to select the plot that exhibited the highest association between the two variables, they might have cued in on the cluster of contaminated data. This will be explored further in Section \ref{sec:TypeIII}.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_prop_correct.pdf}}
       \caption{Proportion of correct responses decreases rapidly with the increase of p-value. After p-value exceeds 0.15 it is very unlikely to identify the actual plot. The theoretical justification of this is shown in Figure \ref{fig:pval_power}. }
       \label{fig:pval_pcorrect}
\end{figure*}

From the experimental data the visual $p$-values are estimated based on Definition \ref{dfn:pvalue}. Figure \ref{fig:pval_definition} displays resulting estimates for each lineup against the conventional $p$-value.  The pattern of visual $p$-values is interesting: for small $p$-values the visual estimates tend to be very small, while lineups with larger $p$-values result in very large visual estimate, giving a clear indication to reject $H_0$ or not. This is why we do not see lot of visual $p$-values  between 0.05 and 0.8 especially for experiment 2. This guides the researcher to make decision confidently while conventional $p$-values take some marginal $p$-values making the decision whether to reject or not harder.

 For visual tests this is not common.

@@

For experiment 3, we see that  the visual $p$ values are very small no matter what the conventional $p$-values are. This is expected since conventional test fails to reject $H_0$ where we should reject, whereas the visual test performs well.

@@

 One assumption made in order to calculate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value. This can be examined for the data collected, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

Figure \ref{fig:P-val_log2} shows the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Columns are ordered according to effect size as defined in section 6.2; rows show replicates for the same combination of parameters.

Red indicates the plot with the lowest $p$-value in the lineup. Blue indicates the plot of the actual data when it is different from that with the lowest $p$-value. In both experiments  people tended to select the plot with the lowest $p$-value with an increase in effect size.
 %This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. 
 The results are clearer for experiment 2, that used a continuous covariate. But even when subjects did not pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

@@

Visual inference falls into the same framework as randomization tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from all of the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between  plots is almost impossible. However,  in this controlled setting we can use $p$-values of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. The detailed discussion is attached to the supplementary material.

@@

A little known error amongst statisticians is what was coined as Type III error in \citet{mosteller:48}. This occurs when the null hypothesis is correctly rejected but for the wrong reason. Experiment 3 is prone to this type of error. Observers were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see any trend. For the human eye this cluster of points is  as visible as the association between the remaining points, leading the observer to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but  still gives the correct answer.

@@

For visual inference, making a Type III, is not actually a problem. It is only a possibility in this experiment because we are working with known structure. In the real setting, we are excited to see observers detecting the actual data plot, and curious about how they detect it, with all possible reasons encapsulated in the alternative hypothesis. This does show the importance of the qualitative answers that observers give for their choices. The detailed discussion on this is attached to the supplementary documents.

@@

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009} as a formal framework to accomplish that. For this, specific terminologies were defined. The methods of obtaining $p$-value and estimating power of visual tests were introduced. If it is assumed that observers tend to pick the plot with the strongest signal, which was consistent with the experimental data, the power of visual test can be obtained theoretically.  Interestingly, the theoretical power of visual test is better than that of conventional test in the conditions tested in this paper. The simulation study showed that even in the situation where the conventional test has its assumptions met, visual inference is comparable. The results also suggest that when the underlying assumptions of a conventional test do not hold, visual inference will perform well. Even though a single observer can evaluate a lineup, as suggested in the initial lineup protocol, this work recommends having multiple observers to ensure enough power.

>>>>

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics.
%These plots should be familiar, because they are all in common use. 
For example, to examine the effect of variable $X_j$ on $Y$, we would plot residuals obtained from fitting the model without $X_j$ against $X_j$ or for a single covariate we may plot $Y$ against $X_j$ (cases 1-4 in Table~\ref{tbl:stat_multiple}). To assess whether the  assumptions of linearity is appropriate we would draw a plot of residuals against fitted values (case 5 in Table~\ref{tbl:stat_multiple}). For the purpose of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable, respectively.

@@

\begin{table*}[hbtp] 
\caption{Visual test statistics for testing hypotheses related to the model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2cm}m{3cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}\end{minipage} & Scatter plot with least square line overlaid. For null plots   we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Residual plot & \begin{minipage}[t]{3cm}   \scalebox{0.4}{\includegraphics{stat_beta_k.pdf}}\end{minipage} & Residual vs $X_k$ plots. For null plots   we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
3 & $H_0: \beta_k=0$ (for binary $X_k$) & Box plot & \begin{minipage}[t]{3cm} 	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For null plots   we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction of continuous and  binary $X_k$) & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_interection.pdf}}  \end{minipage} & Scatter plot with least square lines of each category overlaid. For null plots   we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_nonlinear.pdf}}  \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For null plots   we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{3cm} \scalebox{0.4}{\includegraphics{stat_sigma_box.pdf}}\end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For null plots   we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot of residuals obtained by fitting partial regression. For null plots   we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_goodness_simple.pdf}}  \end{minipage} & Histogram of the response data. For null plots   we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & For  $p=1$ only: $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot with least square line overlaid. For null plots   we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

@@

\begin{equation}\label{power_exp} 
%   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$  increases, and when $\beta$ gets larger.   Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. We observe this dichotomy of visual power  in power estimates based on the data collected from user experiments, too. It features prominently in  figure \ref{fig:power_loess_effect}.

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Three data sets were generated for each of the parameter values shown in Table \ref{tbl:experiment_params} resulting in 60 different ``actual data sets'', and thus, 60 different lineups. For each lineup, the null model was fit to the actual data set to obtain residuals and parameter estimates. The actual data plot was drawn as  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  Figure \ref{fig:test_category} is an example of one of these lineups. It was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. The actual data plot location is ($4^2-1$). For this lineup, 15 out of 16 observers picked the actual data plot.

@@

% Entering 1st row
&  \phantom{0}5 & 0, 1,  3, 5, 8  & 0.25, 0.75, 1.25, 1.75, 2.75 & 0.1, 0.4, 0.75, 1.25, 1.5, 2.25\\[-1ex]
\raisebox{1.5ex}{100} &12
& 1, 3, 8, 10, 16  & 0.5, 1.5, 3.5, 4.5, 6 &\\[1ex]

% Entering 2rd row
&  \phantom{0}5 & 0, 1, 2, 3, 5  & 0.1, 0.4, 0.7, 1, 1.5&\\[-1ex]
\raisebox{1.5ex}{300} & 12
& 1, 3, 5, 7, 10  & 0, 0.8, 1.75, 2.3, 3.5 &\\[1ex]
% [1ex] adds vertical space
\hline
% inserts single-line
\end{tabular}
}
\label{tbl:experiment_params}
\end{table}

@@

This experiment is very similar to the previous one, except that there is a single continuous covariate and no second covariate (Equation \ref{multi} with $p=1$), following the test in Table \ref{tbl:stat_multiple}, case~2. Data is simulated with two sample sizes ($n=100, 300$), two standard deviations of the error ($\sigma=5, 12$), a variety of slopes ($\beta$), as given in Table  \ref{tbl:experiment_params}. We arbitrarily set $\beta_0 = 6$ and values for $X_1$ are simulated from $N(0,1)$. For each combination of parameters, at least three different actual data sets are produced, yielding a total of 70 lineups.

The actual data plot is generated by making a scatterplot of $Y$ vs $X_1$ with the least squares regression line overlaid. To produce the null plots in the lineup null data was simulated from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and plotted using the same scatterplot method as the actual data. 
To select 10 lineups for a subject, each combination of sample size ($n$) and error SD ($\sigma$) is given a difficulty value based on the slope ($\beta$) parameters. For the smallest slopes the difficulty is 4 (hardest) and for the largest slopes the difficulty is 0 (easiest). 
Figure \ref{fig:test_continuous} shows an example lineup for this experiment from difficulty level 4.  This lineup is generated using a sample size ($n$) of 100, slope ($\beta$) of 1.25 and and error SD ($\sigma$) of 5. The actual data plot location is $(2^2+1)$. None of the 65 observers picked the actual plot while 46 observers picked plot 18 which has the lowest $p$-value among all the plots in this lineup.

For each combination of sample size and standard deviation, each participant is given five randomly selected lineups, one of each difficulty level. Another set of four lineups is chosen from a second tier of selected combinations of sample size and standard deviation, with difficulty levels 0 to 3.  A last lineup was randomly selected from a set of  lineups with difficulty level 0. The order in which the lineups are shown to participants is randomized. %Each subject saw a substantial number of difficult lineups.

@@

The first two simulation experiments use data generated under a normal error model, satisfying the conditions for conventional test procedures. In these situations there exists a test, and there would, in general, be no need to use visual inference. The simulation is conducted in the hope that the visual test procedure,  will at least compare favorably with the conventional test -- without any ambition of performing equally well.
 This third simulation is closer to the mark for the purpose of visual inference. The assumptions for the conventional test are violated by contaminating the data. The contamination makes the estimated slopes effectively 0, yet the true value of slope parameter is not.  The data is generated from the following model:

@@

%{\bf Screening scenarios:}
Amazon Mechanical Turk workers are paid for their efforts, not substantially, but on the scale of the minimum wage in the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we use one of the very easy lineups that everybody was exposed to as a {\it reference lineup} and  take action based on a subject's answer to this reference:
if the subject failed to identify the actual data plot on the reference lineup, we remove all of this subject's data from the analysis. If the answer on the reference lineup is correct, we remove the answer for this lineup from the analysis, but keep all of the remaining answers. Table~\ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.  A summary of all the data along with cleaned data are added to the supplementary material.

@@

\subsection{Model fitting}
 For each parameter combination, {\it effect} $E$ is derived as 
%\[
$E=\sqrt {n} \cdot \beta/\sigma.$
%\]
\noindent
The model in Equation \ref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept, i.e. 
 $W_{\ell i} = E_{\ell i}$. Instead of fitting an intercept, we make use of a fixed  offset of $\log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different skill levels of subjects are accounted for  by allowing subject-specific random  slopes for  effect ($E$).

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.55}{\includegraphics{power_loess_effect_k.pdf}}
       \caption{\hh{Make the legend for the lines longer - just as in fig, 3} Power in comparison to effect for the three experiments. Points indicate subject responses, with size indicating count. Responses are 1 and 0 depending on the success or failure respectively to identify the actual plot in the lineup. The loess curve (continuous line) estimates the observed proportion correct (power for $K=1$), and surrounding bands show simultaneous bootstrap confidence band. Observed proportion is used to obtain power for $K=5$. Conventional test power is drawn as a dashed line. For experiment 3, conventional power is based on the slopes of the non-contaminated part of the data. Power of the conventional test for contaminated data is shown by cross marks.}
       \label{fig:power_loess_effect}
\end{figure}

@@

Results of experiment 3 are curious insofar, as  power of the visual test is  largely independent of effect size. However, these results are based on correct identification of the actual data plot, regardless of reason. Although subjects were asked to select the plot that exhibited the highest association between the two variables, they might have cued in on the cluster of contaminated data. This will be explored further in Section \ref{sec:TypeIII}.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_prop_correct.pdf}}
       \caption{Proportion of correct responses decreases rapidly with  increasing $p$-values. For $p$-values above 0.15 it becomes very unlikely that observers identify the actual plot. The theoretical justification of this is shown in Figure \ref{fig:pval_power}. }
       \label{fig:pval_pcorrect}
\end{figure*}

From the experimental data the visual $p$-values are estimated based on Definition \ref{dfn:pvalue}. Figure \ref{fig:pval_definition} displays resulting estimates for each lineup against the conventional $p$-value.  The pattern of visual $p$-values is interesting: for small $p$-values the visual estimates tend to be very small, while lineups with larger $p$-values result in very large visual estimate, giving a clear indication to reject $H_0$ or not. This is why we do not see lot of visual $p$-values  between 0.05 and 0.8 especially for experiment~2. This guides the researcher to make decision confidently while conventional tests with marginal $p$-values make the decision whether to reject or not harder. For visual tests this is not common.

@@

For experiment 3, we see that  the visual $p$ values are very small no matter what the conventional $p$-values are. This is expected as the conventional test loses its power to reject $H_0$ even when the alternative is true, whereas the visual test performs well.

@@

 One assumption made in order to evaluate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value.  We examine  the data collected from the first two experiments, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

\hh{Figure \ref{fig:P-val_log2} gives an overview of all selections in all lineups of experiments 1 and 2. Each panel of the figure corresponds to a single lineup. Each `pin' -- a short line topped by a dot -- corresponds to one plot in the lineup. The $x$-location of the pin shows the plot's $p$-value on a log scale, its height  is given by the number of observer choosing this plot. }
%Figure \ref{fig:P-val_log2} shows the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. 
Columns are ordered according to effect size as defined in section 6.2; rows show replicates for the same combination of parameters.

Red indicates the plot with the lowest $p$-value in the lineup. Blue indicates the plot of the actual data when it is different from that with the lowest $p$-value. In both experiments  people tended to select the plot with the lowest $p$-value. % with an increase in effect size.
 %This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. 
 The results are clearer for experiment 2, that used a continuous covariate. But even when subjects did not pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

@@

Visual inference falls into the same framework as randomization tests, where the statistics from the data are compared with those from null data. Unlike randomization tests visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects' choice is important for understanding the reliability of visual inference. If the actual data plot is very different from all of the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between  plots is almost impossible. However,  in this controlled setting we can use $p$-values of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. The detailed discussion is attached in the supplementary material.

@@

A little known error amongst statisticians is what was coined as Type III error in \citet{mosteller:48}. Type III errors are defined as the probability of  correctly rejecting  the null hypothesis but for the wrong reason. Experiment 3 is prone to this type of error. Participants were asked to identify the plot with the largest absolute slope. But the actual data plot featured a cluster of points, the contamination that made the conventional test  fail to see any trend. For the human eye this cluster of points is  as visible as the association between the remaining points, enabling the observer to identify the actual data plot by looking for the cluster instead of the slope. This would be considered  a Type III error because it leads to a correct rejection of the null hypothesis, but is not related to the value of the slope parameter.

@@

For visual inference, making a Type III, is not actually a problem. It is only a possibility in this experiment because we are working with known structure. In the real setting, we are excited to see observers detecting the actual data plot, and curious about how they detect it, with all possible reasons encapsulated in the alternative hypothesis. However, this highlights the importance of getting qualitative reasoning from observers  for their choices. A more detailed discussion can be found in the supplementary documents.

@@

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009}. Specific terminology was defined, and methods for obtaining the $p$-value and estimating the power of visual tests were introduced. It was assumed that observers select the plot having the strongest signal in the lineup, in order to calculate the theoretical power,  and the experimental data proved to be consistent with this assumption.  The simulation study showed that  results obtained by visual inference are comparable to those obtained by conventional inference, in the setting where %conventional testing can be done. Visual inference is intended to provide valid tests where 
no conventional test exists, and our experiments in a controlled scenario suggest that it will perform as expected in the intended applications. The power of a visual test increases with the number of observers, which interestingly, leads to a result that the theoretical power of visual test can be better than that of conventional tests.
